A Temporal Logic Based Framework for Intrusion Detection

### 2023-09-05 Paper/main.tex
update

According to~\cite{barringer2004rule}, the definitions of $evaluate$, $update$ and $value$ on the different primitive operators are the following ($op\text{ can be }\lor , \land, \rightarrow $):
%evaluate
\begin{align*}
    evaluate(\underline{\text{true}},\alpha_i) & =\ \underline{\text{true}} \\
    evaluate(\underline{\text{false}},\alpha_i) & =\ \underline{\text{false}} \\ 
    evaluate(exp,\alpha_i) & =\ \text{value of } exp\text{ in } \alpha_i \\
    evaluate(F_1\ op\ F_2,\alpha_i) & =\  evaluate(F_1,\alpha_i)\ op\ evaluate(F_2,\alpha_i)\\
    evaluate(\neg{F},\alpha_i) & =\ \neg{evaluate(F,\alpha_i)}  \\
    evaluate(\bigcirc F,\alpha_i) & =\ update(F,\alpha_i)  \\
    evaluate(F_1 \cdot F_2,\alpha_i) & =
    \begin{cases}
        evaluate(F_1,\alpha_i)\cdot F_2 \quad \text{if } value(F_1)=\ \underline{\text{false}} \\
        evaluate(F_1,\alpha_i)\cdot F_2 \lor evaluate(F_2,\alpha_i)\quad \text{else} 
    \end{cases} \\ \\
%\end{align*}
%\noindent\begin{minipage}[l]{\textwidth}
   % \raggedright
    %\begin{align*}
        \text{value}(\underline{\text{true}}) & = \underline{\text{true}} \\
        \text{value}(\underline{\text{false}}) & = \underline{\text{false}} \\ 
        \text{value}(\text{exp}) & = \underline{\text{false}} \\
        \text{value}(F_1\ \text{op}\ F_2) & = \text{value}(F_1)\ \text{op}\ \text{value}(F_2) \\
        \text{value}(\neg{F}) & = \neg{\text{value}(F)}  \\
        \text{value}(\bigcirc F) & =
        \begin{cases}
            F & \text{if at the beginning of a trace} \\
            \underline{\text{false}} & \text{if at the end of a trace} \\
        \end{cases}\\
        \text{value}(F_1 \cdot F_2,\alpha_i) & = \text{value}(F_1)\land \text{value}(F_2) \\ \\
   % \end{align*}
%\end{minipage}
%\hfill
%\begin{minipage}[r]{\textwidth}
   % \raggedleft
    %\begin{align*}
        \text{update}(\underline{\text{true}},\alpha_i) & = \underline{\text{true}} \\
        \text{update}(\underline{\text{false}},\alpha_i) & = \underline{\text{false}} \\ 
        \text{update}(\text{exp},\alpha_i) & = \text{exp}\\
        \text{update}(F_1\ \text{op}\ F_2,\alpha_i) & = \text{update}(F_1,\alpha_i)\ \text{op}\ \text{update}(F_2,\alpha_i) \\
        \text{update}(\neg{F},\alpha_i) & = \neg{\text{update}(F,\alpha_i)}  \\
        \text{update}(\bigcirc F,\alpha_i) & = \text{update}(F,\alpha_i)  \\
        \text{update}(F_1 \cdot F_2,\alpha_i) & = \text{update}(F_1,\alpha_i)\cdot F_2
    \end{align*}
%\end{minipage}
Let us introduce an example to better understand the algorithm. Given the following rules 
\begin{align*}
    &\underline{\text{max}}\ \text{Always}(\underline{\text{Form}}\ F)  = F \land \bigcirc \text{Always}(F) \\
    &\underline{\text{min}}\ \text{EvTimedLogout}(\underline{\text{string }} k, \underline{\text{double }} t, \underline{\text{double }} \delta) = (\text{time} - t \leq \delta) \\
    &\quad\quad\quad\quad \land ((action = \text{logout} \land userId = k) \lor \bigcirc \text{EvTimedLogout}(k, t, \delta)) \\
    &\underline{\text{mon}}\ M_3  = \text{Always}((action = \text{login}) \rightarrow \text{EvTimedLogout}(userId,time, 100))
\end{align*}
and a trace $\sigma=\alpha_1,\alpha_2$, where $\alpha_1=\{userId:\text{"Lori"},action:\text{login},time:\text{17.0}\}$ and $\alpha_1=\{userId:\text{"Lori"},action:\text{logout},time:\text{150.0}\}$, the initial formula $M_3$ gets modified. At time $1$, when $\alpha_1$ is analyzed, $M_3$ changes as follows:
\begin{align*}
    &evaluate(M_3,\alpha_1)  = \text{EvTimedLogout("Lori",17.0, 100)} \land \\ 
    &\quad\quad\quad\quad \text{Always}((action = \text{login}) \rightarrow \text{EvTimedLogout}(userId,time, 100))
\end{align*}
At the second event the predicate $(time-t \leq \delta)$ gets instantiated to $(150.0-17.0 \leq 100)$ which is false; thus the whole formula becomes false, indicating that the two-event trace violates the property $M_3$.

It's essential to emphasize that the logic EAGLE operates with a \textit{finite} trace semantics; the monitoring algorithm, as written before, can check whetet the formula is satisfied or not only at the conclusion of a trace. However, in intrusion detection scenarios, the notion of the 'end of a trace' makes no sense since event sequences can theoretically be infinite. In such cases, the aim is to trigger an alarm as soon as a property is violated and this is done by continuously checking the formula's satisfaction status after each event.

>>>>

\textbf{Calculus.} According to~\cite{barringer2004rule}, the definitions of $evaluate$, $update$ and $value$ on the different primitive operators are the following ($op\text{ can be }\lor , \land, \rightarrow $):
%evaluate
\begin{align*}
    evaluate(\underline{\text{true}},\alpha_i) & =\ \underline{\text{true}} \\
    evaluate(\underline{\text{false}},\alpha_i) & =\ \underline{\text{false}} \\ 
    evaluate(exp,\alpha_i) & =\ \text{value of } exp\text{ in } \alpha_i \\
    evaluate(F_1\ op\ F_2,\alpha_i) & =\  evaluate(F_1,\alpha_i)\ op\ evaluate(F_2,\alpha_i)\\
    evaluate(\neg{F},\alpha_i) & =\ \neg{evaluate(F,\alpha_i)}  \\
    evaluate(\bigcirc F,\alpha_i) & =\ update(F,\alpha_i)  \\
    evaluate(F_1 \cdot F_2,\alpha_i) & =
    \begin{cases}
        evaluate(F_1,\alpha_i)\cdot F_2 \quad \text{if } value(F_1)=\ \underline{\text{false}} \\
        evaluate(F_1,\alpha_i)\cdot F_2 \lor evaluate(F_2,\alpha_i)\quad \text{else} 
    \end{cases} \\ \\
%\end{align*}
%\noindent\begin{minipage}[l]{\textwidth}
   % \raggedright
    %\begin{align*}
        \text{value}(\underline{\text{true}}) & = \underline{\text{true}} \\
        \text{value}(\underline{\text{false}}) & = \underline{\text{false}} \\ 
        \text{value}(\text{exp}) & = \underline{\text{false}} \\
        \text{value}(F_1\ \text{op}\ F_2) & = \text{value}(F_1)\ \text{op}\ \text{value}(F_2) \\
        \text{value}(\neg{F}) & = \neg{\text{value}(F)}  \\
        \text{value}(\bigcirc F) & =
        \begin{cases}
            F & \text{if at the beginning of a trace} \\
            \underline{\text{false}} & \text{if at the end of a trace} \\
        \end{cases}\\
        \text{value}(F_1 \cdot F_2,\alpha_i) & = \text{value}(F_1)\land \text{value}(F_2) \\ \\
   % \end{align*}
%\end{minipage}
%\hfill
%\begin{minipage}[r]{\textwidth}
   % \raggedleft
    %\begin{align*}
        \text{update}(\underline{\text{true}},\alpha_i) & = \underline{\text{true}} \\
        \text{update}(\underline{\text{false}},\alpha_i) & = \underline{\text{false}} \\ 
        \text{update}(\text{exp},\alpha_i) & = \text{exp}\\
        \text{update}(F_1\ \text{op}\ F_2,\alpha_i) & = \text{update}(F_1,\alpha_i)\ \text{op}\ \text{update}(F_2,\alpha_i) \\
        \text{update}(\neg{F},\alpha_i) & = \neg{\text{update}(F,\alpha_i)}  \\
        \text{update}(\bigcirc F,\alpha_i) & = \text{update}(F,\alpha_i)  \\
        \text{update}(F_1 \cdot F_2,\alpha_i) & = \text{update}(F_1,\alpha_i)\cdot F_2
    \end{align*}
%\end{minipage}
Let us introduce an example to better understand the algorithm. Given the following rules 
\begin{align*}
    &\underline{\text{max}}\ \text{Always}(\underline{\text{Form}}\ F)  = F \land \bigcirc \text{Always}(F) \\
    &\underline{\text{min}}\ \text{EvTimedLogout}(\underline{\text{string }} k, \underline{\text{double }} t, \underline{\text{double }} \delta) = (\text{time} - t \leq \delta) \\
    &\quad\quad\quad\quad \land ((action = \text{logout} \land userId = k) \lor \bigcirc \text{EvTimedLogout}(k, t, \delta)) \\
    &\underline{\text{mon}}\ M_3  = \text{Always}((action = \text{login}) \rightarrow \text{EvTimedLogout}(userId,time, 100))
\end{align*}
and a trace $\sigma=\alpha_1,\alpha_2$, where $\alpha_1=\{userId:\text{"Lori"},action:\text{login},time:\text{17.0}\}$ and $\alpha_1=\{userId:\text{"Lori"},action:\text{logout},time:\text{150.0}\}$, the initial formula $M_3$ gets modified. At time $1$, when $\alpha_1$ is analyzed, $M_3$ changes as follows:
\begin{align*}
    &evaluate(M_3,\alpha_1)  = \text{EvTimedLogout("Lori",17.0, 100)} \land \\ 
    &\quad\quad\quad\quad \text{Always}((action = \text{login}) \rightarrow \text{EvTimedLogout}(userId,time, 100))
\end{align*}
At the second event the predicate $(time-t \leq \delta)$ gets instantiated to $(150.0-17.0 \leq 100)$ which is false; thus the whole formula becomes false, indicating that the two-event trace violates the property $M_3$.

It's essential to emphasize that the logic EAGLE operates with a \textit{finite} trace semantics; the monitoring algorithm, as written before, can check whetet the formula is satisfied or not only at the conclusion of a trace. However, in intrusion detection scenarios, the notion of the 'end of a trace' makes no sense since event sequences can theoretically be infinite. In such cases, the aim is to trigger an alarm as soon as a property is violated and this is done by continuously checking the formula's satisfaction status after each event. In practice, once we have the specification of the attack pattern in terms of $\varphi$, we can express the system's safe behavior as $\square(\neg \varphi)$. Note that if a sequence of events corresponding to the attack is discovered, this formula becomes false.
